\vspace{-0.5in}{\bfseries\scshape Transfer Learning for computationally limited training of Vision Transformers on art classification tasks } \\ \vspace{1ex} \normalsize{Bachelor's Project Thesis}\vspace{-2ex}

### 2022-06-13 writing/tex/draft_methods.tex
Worked on the results section

% Experiments build upon Sabatelli et al.
Much of the experiments and analysis discussed below, builds upon the work done in \citeauthor{sabatelli2018deep} (\citeyear{sabatelli2018deep}). This paper investigated TL properties of CNNs in the art classification domain. The main focus was on a comparison between CNN architectures pretrained on ImageNet1K when they were either used as OTS feature extractors, or fine-tuned to the new task. It was suggested that either approach was beneficial compared to training from scratch, but that FT resulted in substantially better performance than OTS TL.

@@

\subsection{Models}
Eight different neural network architectures are used throughout this paper, of which four are CNN-based and four ViT-based. When multiple versions are available (i.e. `base', `small', `tiny'), `base'/`medium' ones are chosen.

To start off with CNNs, ResNet50 \citep{he2016deep} and VGG19 \citep{simonyan2014very} are chosen, as these were the best performing models in \citeauthor{sabatelli2018deep} for FT an OTS TL, respectively. In addition, \citeauthor{matsoukas2021time} (\citeyear{matsoukas2021time}) used ResNet50 too, while \citeauthor{zhou2021convnets} (\citeyear{zhou2021convnets}) used ResNet101 and ResNet152. To also add some more recent CNNs, ConvNext \citep{liu2022convnet} -- a purely CNN-based model inspired by ViTs' recent success -- and EfficientNetV2 \citep{tan2021efficientnetv2} are included as well.

For ViTs, versions with a 16x16 patch size are chosen. The first model is the original ViT \citep{dosovitskiy2020image}. Swin \citep{liu2021swin}, which was mentioned in section \ref{related_work}, is also selected, as it showed promising performance in \citeauthor{zhou2021convnets}. DeiT \citep{touvron2021training}, the distillation based architecture, was also mentioned in section \ref{related_work}. The small version of this architecture was used in \citeauthor{matsoukas2021time}, as it is similar in size to ResNet50, to which it was compared. Here, however, a base-sized version is chosen. Specifically, the version that does not include output for distillation learning.
% TODO if you do end up using distilation with ConvNext as a teacher: don't forget to change this section!!!

All ViT-based models have a bit over 86 million parameters. ConvNext is of similar size, with 88.6 million parameters. ResNet50 and EfficientNetV2 are smaller, with 25.6 and 13.6 million respectively, while VGG19 is the largest model, with 143.7 parameters. It should be mentioned that for both OTS TL as FT, preliminary experiments were performed with `small' and `tiny' models. The results of these suggested that similar conclusion could have been drawn if these were used instead.

@@

For all models, the linear classification layer is replaced as to fit the new task. Cross entropy loss uses the raw outputs of this layer (conform PyTorch documentation), together with one-hot encoded prediction targets.

@@

\subsection{Hardware and software}
All experiments are conducted on a single compute node containing one 32GB Nvidia V100 GPU. These nodes are constituents of the \textit{Peregrine high performance computing cluster}, belonging to the \textit{University of Groningen's Center for Information Technology}. The FT experiments take advantage of the V100's mixed precision acceleration. An exception is made for the type classification experiment, as this one is also used to compare OTS TL and FT in terms of time/accuracy trade-offs. Preliminary findings suggested that using mixed precision does not affect performance for the experiments where it is used.

>>>>

% Experiments build upon Sabatelli et al.
Much of the experiments and analysis discussed below, builds upon the work done in \citeauthor{sabatelli2018deep} (\citeyear{sabatelli2018deep}). This paper investigated TL properties of CNNs in the art classification domain. The main focus was on a comparison between CNN architectures pretrained on ImageNet1K when they were either used as OTS feature extractors, or fine-tuned to the new task. It was suggested that FT resulted in substantially better performance than training from scratch or OTS TL.

@@

\subsection{Models} \label{methods:models}
Eight different neural network architectures are used throughout this paper, of which four are CNN-based and four ViT-based. When multiple versions are available (i.e. `base', `small', `tiny'), `base'/`medium' ones are chosen.

To start off with CNNs, ResNet50 \citep{he2016deep} and VGG19 \citep{simonyan2014very} are chosen, as these were the best performing models in \citeauthor{sabatelli2018deep} for respectively FT and OTS TL. In addition, \citeauthor{matsoukas2021time} (\citeyear{matsoukas2021time}) used ResNet50 too, while \citeauthor{zhou2021convnets} (\citeyear{zhou2021convnets}) used ResNet101 and ResNet152. To also add some more recent CNNs, ConvNext \citep{liu2022convnet} -- a purely CNN-based model inspired by ViTs' recent success -- and EfficientNetV2 \citep{tan2021efficientnetv2} are included as well.

For ViTs, versions with a 16x16 patch size are chosen. The first model is the original ViT \citep{dosovitskiy2020image}. Swin \citep{liu2021swin}, which was mentioned in section \ref{related_work}, is also selected, as it showed promising performance in \citeauthor{zhou2021convnets}. DeiT \citep{touvron2021training}, the distillation based architecture, was also mentioned in section \ref{related_work}. The small version of this architecture was used in \citeauthor{matsoukas2021time}, as it is similar in size to ResNet50, to which it was compared. Here, however, also a base-sized version is chosen. Specifically, the version that does not include output for distillation learning.
% TODO if you do end up using distilation with ConvNext as a teacher: don't forget to change this section!!!

All ViT-based models have a bit over 86 million parameters. ConvNext is of similar size, with 88.6 million parameters. ResNet50 and EfficientNetV2 are smaller, with 25.6 and 13.6 million, respectively, while VGG19 is the largest model, with 143.7 million parameters. It should be mentioned that for both OTS TL and FT, preliminary experiments were performed with `small' and `tiny' models. The results of these suggested that similar conclusions could have been drawn if these were used instead.

@@

For all models, the linear classification layer is replaced as to fit the new task. Cross entropy loss uses the raw outputs of this layer (conform PyTorch documentation), together with one-hot encoded prediction targets. An early stop occurs after 10 epochs without improved loss on the validation set, and the model with the lowest loss is benchmarked on the testing set.

@@

\subsection{Hardware and software}
All experiments are conducted on a single compute node containing one 32 GB Nvidia V100 GPU. These nodes are constituents of the \textit{Peregrine high performance computing cluster}, belonging to the \textit{University of Groningen's Center for Information Technology}. The FT experiments take advantage of the V100's mixed precision acceleration. An exception is made for the type classification experiment, as this one is also used to compare OTS TL and FT in terms of time/accuracy trade-offs. Preliminary findings suggested that using mixed precision does not affect performance for the aforementioned experiments.

### 2022-06-13 writing/tex/draft_results.tex
Sort of finished the first draft of the Methods section. Probably will still change some thing before I move on to results, however..
Worked on the results section

\begin{figure*}
    \centering
    \begin{subfigure}{0.32\textwidth}
    \def\svgwidth{5.5cm}
    \input{img/ots_type_acc.pdf_tex}
    \caption{Type}
    \label{results:img:ots_type}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.32\textwidth}
    \def\svgwidth{5.5cm}
    \input{img/ots_mat_acc.pdf_tex}
    \caption{Material}
    \label{results:img:ots_mat}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.32\textwidth}
    \def\svgwidth{5.5cm}
    \input{img/ots_artist_acc.pdf_tex}
    \caption{Artist}
    \label{results:img:ots_artist}
    \end{subfigure}
    \caption{Validation accuracy using off the shelf models.}
\end{figure*}

@@

\begin{figure*}
    \centering
    \begin{subfigure}{0.32\textwidth}
    \def\svgwidth{5.5cm}
    \input{img/ft_type_acc.pdf_tex}
    \caption{Type}
    \label{results:img:ft_type}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.32\textwidth}
    \def\svgwidth{5.5cm}
    \input{img/ft_mat_acc.pdf_tex}
    \caption{Material}
    \label{results:img:ft_mat}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.32\textwidth}
    \def\svgwidth{5.5cm}
    \input{img/ft_artist_acc.pdf_tex}
    \caption{Artist}
    \label{results:img:ft_artist}
    \end{subfigure}
    \caption{Validation accuracy using fine-tuned models.}
\end{figure*}

@@

@@

@@

>>>>

This section presents results of the conducted experiments, mostly in the form of tables and line plots. Tables show average testing performance over the 5 trials per experiment, with standard deviation ($s$) between brackets. A green shaded cell marks best performance, while yellow and red depict second best and worst performance, respectively.

Line plots show average validation accuracy (dashed line is CNN-based, continuous is ViT-based), with \textit{standard error of the mean} ($\pm s \div \sqrt{N}$, where $N=5$) depicted as a shaded region. Plots end when an early stop occurred for the first of 5 trials.

% Tell that the appendix shows all other plots as well maybe. Only for the BP paper tho.

\subsection{Off the shelf learning}

\begin{figure*}
    \centering
    \begin{subfigure}{0.32\textwidth}
    \def\svgwidth{5.5cm}
    \input{img/ots_type_acc.pdf_tex}
    \caption{Type}
    \label{results:img:ots_type}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.32\textwidth}
    \def\svgwidth{5.5cm}
    \input{img/ots_mat_acc.pdf_tex}
    \caption{Material}
    \label{results:img:ots_mat}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.32\textwidth}
    \def\svgwidth{5.5cm}
    \input{img/ots_artist_acc.pdf_tex}
    \caption{Artist}
    \label{results:img:ots_artist}
    \end{subfigure}
    \caption{Validation accuracy using off the shelf models.}
    \label{results:img:ots}
\end{figure*}

@@

Tables \ref{results:tab:ots_type} through \ref{results:tab:ots_artist} show performance of OTS trained models on the 3 classification tasks, while figure \ref{results:img:ots} shows corresponding validation accuracy per epoch. Performance on the testing set is comparable on all tasks in terms of ranking. Other observations are consistent as well, such as ConvNext taking the longest to converge.

Swin is coming out on top in all cases, which is in line with results of \citeauthor{zhou2021convnets} (\citeyear{zhou2021convnets}). What slightly deviates from these result, is the observation that ViT performs better than a ResNet-based model on 2 out of 3 tasks. This comparison is, however, not very meaningful, as \citeauthor{zhou2021convnets} uses ResNet101 and 152 instead of ResNet50.

It is interesting that VGG19 performs worse than ResNet50, as it was the best performing OTS model on all 3 tasks in \citeauthor{sabatelli2018deep} (\citeyear{sabatelli2018deep}). This discrepancy is likely attributable to the difference in dataset size, as VGG19 is by far the largest model (see section \ref{methods:models}), and likely too large for such small datasets.

When comparing the tasks to one another, it can be observed that overall accuracies are lowest for material classification, and highest for artist classification. This seems counterintuitive, as one would think artist identification is much more challenging than determining if something is a painting, or if it is made out of paper instead of wood. An explanation can likely be found in table \ref{methods:datasets}, which shows that material and artist classification sets are the least and most balanced, respectively. In addition, tables \ref{results:tab:ots_type} through \ref{results:tab:ots_artist} show that balanced accuracies are up to 14\% lower than plain ones for material classification, while they are much closer for artist classification. This suggests much accuracy is lost by misclassifying instances of less occurring classes in favor of more occurring ones. Similar finding can be done for FT results in section \ref{results:ft}.
% Maybe switch this to discussion tho ^

%OTS TL:
%    Comparisons with Sabatelli paper (ResNet50 vs VGG19)
%    ViTs slightly better
%    Lower balanced accuracy
%    Higher balanced accuracy for type classification and even more for artist
%    Artist higherst, which is surprising since it is the most difficult intuitively

\subsection{Fine-tuning} \label{results:ft}

\begin{figure*}
    \centering
    \begin{subfigure}{0.32\textwidth}
    \def\svgwidth{5.5cm}
    \input{img/ft_type_acc.pdf_tex}
    \caption{Type}
    \label{results:img:ft_type}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.32\textwidth}
    \def\svgwidth{5.5cm}
    \input{img/ft_mat_acc.pdf_tex}
    \caption{Material}
    \label{results:img:ft_mat}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.32\textwidth}
    \def\svgwidth{5.5cm}
    \input{img/ft_artist_acc.pdf_tex}
    \caption{Artist}
    \label{results:img:ft_artist}
    \end{subfigure}
    \caption{Validation accuracy using fine-tuned models.}
    \label{results:img:ft}
\end{figure*}

@@

\subsection{Scaling dataset size}

@@

\subsection{Off the shelf learning versus fine-tuning}

@@

\subsection{Qualitative analysis}
TODO
